### 2020-11-12 manuscript/manuscript.rmd
update acknowledgement
Updated model description

The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories. ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data. Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes]. It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses. To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa matrices, we conducted Bayesian cross-validation experiments with PhyloBayes. We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$). This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model. This suggests there is no reason to favor the narrow analysis conditions under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some such analyses. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency (Fig. 4). The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though. This suggests that we shouldn't privilege these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.

Pin-pointing category number as an issue with large effect on analyses of the animal root will help guide future analyses that address this question. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and that this problem can be a test-bed to develop methods and tools that will help with other difficult phylogenetic problems as well. Advances on the question of the animal root will come from progress on other fronts as well. For example, there are many organisms that are highly relevant to this problem, in particular outgroup, ctenophore, and sponge taxa, for which no genome or transcriptome data are available [@King:2017ie]. More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria. It will be critical for this to change in coming years.

@@

We thank members of the Dunn and Rokas laboratories for discussions and comments. We are grateful to Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript. We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster. Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad. This work was conducted in part using the resources of the Advanced Computing Center for Research and Education (ACCRE) at Vanderbilt University. Research in A.R.'s lab is supported by the National Science Foundation (DEB-1442113), the National Institutes of Health / National Institute of Allergy and Infectious Diseases (1R56AI146096-01A1), the Guggenheim Foundation, and the Burroughs Wellcome Fund.

>>>>

The first questions regard model fit. ModelFinder selects site heterogeneous C60 models according to BIC (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). In PhylobBayes, Cross-validation is a reliable and suggested approach to model evaluation [@lartillot2009phylobayes]. We elvaluated  Poisson+CAT and Poisson+nCAT60 models with cross-validation in PhyloBayes for the Whelan2017_strict and Philippe2009_Choanozoa matrices. For both matrices, we found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$). Cross-validation therefore does not indicate that the unconstrained CAT models are a better fit than the models with 60 categories.  This suggests there is not a reason to favor the narrow analysis conditions (Choanozoa taxon sampling, Poisson exchange matrix, and unconstrained CAT models that use hundreds of categories) under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data may help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some such analyses. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency (Fig. 4). The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model fit according to cross-validation, though. This suggests that we shouldn't privilege these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.

Pin-pointing category number as an issue with large effect on analyses of the animal root will help guide future analyses that address this question. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and that this problem can be a test-bed to develop methods and tools that will help with other difficult phylogenetic problems as well. Advances on the question of the animal root will come from progress on other fronts as well. For example, there are many organisms that are highly relevant to this problem, in particular outgroup, ctenophore, and sponge taxa, for which no genome or transcriptome data are available [@King:2017ie]. More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria. Future analyses focused on complete genomes rather than transcriptomes and partial genomes will have multiple advantages. Data matrices derived from these more complete sources will have a lower fraction of missing sequences. Complete gene sampling within each species will also greatly improve analyses of gene duplciation and loss, a crititical step in building phylogenomic matrices such as those presented here. Analyses of this new generation of matrices derived from complete genomes will be well served my understandind the sources of analysis variation in the generation of matrices that came before them.

@@

We thank members of the Dunn and Rokas laboratories for discussions and comments. We are grateful to Steve Haddock, Kyle David, and Nathan Whelan for feedback on an earlier version of this manuscript. We are grateful to Nicole King, Daniel Richter, Paul Lewis, Warren Francis and Joseph Ryan for feedback on the preprint version of this manuscript. We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster. Yuanning Li was partially supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad. This work was conducted in part using the resources of the Advanced Computing Center for Research and Education (ACCRE) at Vanderbilt University. Research in A.R.'s lab is supported by the National Science Foundation (DEB-1442113), the National Institutes of Health / National Institute of Allergy and Infectious Diseases (1R56AI146096-01A1), the Guggenheim Foundation, and the Burroughs Wellcome Fund.
